Quantifying the importance of an inference model in Bayesian phylogenetics

### 2020-02-11 supplement.tex
Weave a better story at the supplementary materials

First, subsection \ref{subsec:main_example} shows extra figures and
 diagnostics regarding the main example. Because the main example
only uses one exemplary tree, subsection \ref{subsec:distribution}
shows the result of using multiple trees, as generated by the
same stochastic process as the main example.

@@

\richel{Issue \url{https://github.com/richelbilderbeek/pirouette_article/issues/75}}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{The effect of the number of taxa}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{The effect of DNA sequence length}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{The effect of non-clock like models}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

The code used in this part of the article can be found at 
\url{https://github.com/richelbilderbeek/pirouette_example_3007}.

\begin{figure}[H]
  \includegraphics[width=\textwidth]{pirouette_example_3007/example_17_314/errors.png}
  \caption{An unliked node substitution model}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{The effect of assuming a Yule tree prior on a Yule tree}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{The effect of assuming a Yule tree prior on a BD tree}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{The effect of differently common diversity-dependent trees}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Here we show the results of a \verb;pirouette; run on a dataset of $N = 100$ \giovanni{N can be changed if 100 is too much} trees simulated under a diversity-dependent process. All the trees are simulated using the DDD package (\cite{DDD}). Our goal is to show how trees simulated by a process that is sensibly different from a standard tree prior will result in a significant divergence between true and twin distribution. To do so we set the simulation parameters (in this case: absolute speciation rate $\lambda_0 = 1.6$, extinction rate $\mu = 0.2$ and carrying capacity $K = 20$) in order to express clear characteristics of diversity dependence in the phylogenies. In fact, for a high speciation rate and a low carrying capacity we expect trees to exhibit high unbalance, compared with trees generated by a standard birth-death model. We can measure how unbalanced is a tree using the $\gamma$ statistics (\cite{pybus2000testing}). For this reason, we choose to measure the error made by the inference process using the $\gamma$ error statistics, which is currently implemented in \verb;pirouette;.

@@

\begin{figure}[H]
  \includegraphics[width=\textwidth]{pirouette_example_3008/example_18_314/errors.png}
  \caption{Equal mutation rate}
\end{figure}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Overview}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

For theoreticians:

\itemize{
  \item subsection \ref{subsec:main_example}: extra figures and
    diagnostics regarding the main example.
  \item subsection \ref{subsec:distribution}: the result of using 
    multiple trees, as generated by the
    same stochastic process as the main example
}

For empiricists:

\itemize{
  \item subsection \ref{subsec:n_taxa}: the effect of the number
    of taxa
  \item subsection \ref{subsec:n_nucleotides}: the effect of the DNA
    alignment sequence length
  \item subsection \ref{subsec:non_clock}: the effect when the alignment
    is generated by a non-clock like nucleotide substitution model
  \item subsection \ref{subsec:simplest_correct_parameterization} shows the
    effect when doing an inference in the simplest use case
  \item subsection \ref{subsec:under_parameterization} shows the
    effect when doing an inference with an under-parameterization
  \item subsection \ref{subsec:better_label_needed} shows 
    \richel{better description needed}

}

@@

\richel{Issue at \url{https://github.com/richelbilderbeek/pirouette_article/issues/75}}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{The effect of the number of taxa}
\label{subsec:n_taxa}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\richel{Issue at \url{https://github.com/richelbilderbeek/pirouette_article/issues/63}}

The main example uses \richel{?} taxa.
One expects that the more taxa are present in a phylogeny,
the better the inference will be, 
thus the lower the errors will be.
Here, we show the same results as the main example,
except for a varying number of taxa.

Note that, \richel{[some authors]} found that an increase in the number
of taxa may counterintuitively decrease the inference.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{The effect of DNA sequence length}
\label{subsec:n_nucleotides}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\richel{Issue at \url{https://github.com/richelbilderbeek/pirouette_article/issues/64}}

The main example uses a DNA alignment length of \richel{?} nucleotides.
One expects that the more nucleotides are present in the alignment,
the more information is available,
thus the better the inference will be, 
thus the lower the errors will be.
Here, we show the same results as the main example,
except for a varying DNA alignment sequence length.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{The effect of non-clock like models}
\label{subsec:non_clock}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\richel{Issue at \url{https://github.com/richelbilderbeek/pirouette_article/issues/87}}

The main example generates an alignment using a strict clock model (that is,
all species have an equal mutation rate).
One may wonder, how well the inference will be for a different non-clock like
model.
Here, we show the same results as the main example,
except that instead of generating the alignment with a strict clock
model, we use an unlinked node substitution model (by Thijs Janzen,
unpublished, code available at \url{https://github.com/thijsjanzen/nodeSub}),
with a branch mutation rate of $0.1$ and a node mutation rate of $0.1$.

The code used in this part of the article can be found at 
\url{https://github.com/richelbilderbeek/pirouette_example_17}.

\begin{figure}[H]
  \includegraphics[width=\textwidth]{pirouette_example_17/example_17_314/errors.png}
  \caption{An unliked node substitution model}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{The effect of assuming a Yule tree prior on a Yule tree}
\label{subsec:simplest_correct_parameterization}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\richel{Issue at \url{https://github.com/richelbilderbeek/pirouette_article/issues/66}}

The main example uses a tree generated by a non-standard tree model.
Here, we show the same results, with the only difference that
the tree used is generated by simplest tree model (the Yule model),
which we also assume as the (correct) tree prior.
This example thus shows a parameterization at the correct level for the
simplest case possible.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{The effect of assuming a Yule tree prior on a BD tree}
\label{subsec:under_parameterization}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\richel{Issue at \url{https://github.com/richelbilderbeek/pirouette_article/issues/67}}

The main example uses a tree generated by a non-standard tree model.

Here, we show the same results, with the difference that
the tree used is generated by a birth-death (BD) tree model,
where we assume it is generated by a Yule (or pure-birth) model.
This example thus shows the effect of underparameterization.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{The effect of differently common diversity-dependent trees}
\label{subsec:better_label_needed}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\richel{Issue at \url{https://github.com/richelbilderbeek/pirouette_article/issues/68}}

Here we show the results of a \verb;pirouette; run on a dataset 
of $N = 100$ \giovanni{N can be changed if 100 is too much} 
trees simulated under a diversity-dependent process. 
All the trees are simulated using the DDD package (\cite{DDD}). 
Our goal is to show how trees simulated by a process that is 
sensibly different from a standard tree prior will result in a 
significant divergence between true and twin distribution. 
To do so we set the simulation parameters (in this case: 
absolute speciation rate $\lambda_0 = 1.6$, 
extinction rate $\mu = 0.2$ and carrying capacity $K = 20$) 
in order to express clear characteristics of diversity dependence in the 
phylogenies. In fact, for a high speciation rate and a low carrying capacity, 
we expect trees to exhibit high unbalance, compared with trees generated 
by a standard birth-death model. 
We can measure how unbalanced is a tree using the $\gamma$ 
statistics (\cite{pybus2000testing}). For this reason, we choose to 
measure the error made by the inference process using the $\gamma$ error statistics, 
which is currently implemented in \verb;pirouette;.

@@

\begin{figure}[H]
  \includegraphics[width=\textwidth]{pirouette_example_18/example_18_314/errors.png}
  \caption{Equal mutation rate}
\end{figure}
